A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-26 paper.tex
Minor edits

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees in Fig.~\ref{fig-simplification}B. The operation to remove them,
the results of which are shown in Fig.~\ref{fig-simplification}C,
therefore requires knowledge not just of the graph topology, but also of the
edge annotations. Removal of recombinant nodes can produce graph nodes with
more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node,
something which is impossible to encode in an eARG.

@@

We have seen that the local tree for a given position on
the genome can be extracted from an ARG by starting at the
leaves and following the appropriate paths through the graph.
However, this tells us little about the relationship \emph{between}
adjacent trees, and what properties the overall emsemble of
local trees might have.
A recombination between two adjacent trees causes them to
differ by a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the ``subtree transfer'' operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

The effects of recombination as we move from left-to-right along
the genome in an unsimplified gARG are easy to interpret.
Consider the highlighed tree in Fig~\ref{fig-simplification}A.
In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
its parent is \textsf{k}. Since both parents are present as
unary nodes in the tree, we simply change one edge in
order to effect the transition.
% Need revising to fit new example
A more complex transition occurs when we have a regraft operation
that occurs above the current root (i.e., the current oldest
non-unary node). For example, the transition between the
second and third trees corresponds to node D switching
from the path through I to H, and we then must add branches
corresponding to passing through L and meeting at M.

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is important to note that while extracting local
trees from an ARG is straightforward, the converse
process of reconstructing precisely the same ARG from
these trees is more problematic. If the internal
nodes are consistently labelled and unary nodes are
included in the local trees (e.g., Fig~\ref{fig-simplification})
then it is easy to exactly reconstruct the ARG.
Otherwise, the problem is much more difficult and it
may not be possible to uniquely reconstruct the original ARG.
Without labels for the internal nodes, we are forced to
to reason about which subtrees remain the same and which
differ due to the SPR separating adjacent trees in order to
assign node identity. Without the internal unary nodes marking
recombinations, we must infer their existence. In
general, there will not be a unique solution to this problem,
and so there can be many possible ARGs compatible with
a given list of leaf-labelled binary trees.
This is the approach taken by the Espalier inference
method~\citep{rasmussen2022espalier}, which finds a
set of SPRs that reconcile independently inferred
trees from different regions of the genome
using.

\section*{Precision of ancestry information}
Ancestral recombination graphs are often characterised
as consisting of a complete record of the genealogical history
of a sample. For example,
\cite{rasmussen2014genome} state that an ``ARG provides a record of all
coalescence and recombination events since the divergence of the sequences
under study'' and
according to \cite{deng2021distribution} an ARG
``provides all the information about the genealogical history of a sample,
including the locations of recombination events.'' It is worth
questioning, however, whether such comprehensive information is
realistically available, and in particular whether our
representation of genealogical history should \emph{require}
such precision.

@@

% JK: This is very rough, just jotting down the basic ideas,
% will need substantial revision.
It is not clear that such complete information will always
(or indeed ever) be available from inferences. We may be
able to precisely identify the location and timing
of recent recombinations (particularly if detailed pedigree
information is available) but beyond this, the information
required simply isn't present. When sampling from a
process such as the Sequentially Markov
Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
like ARGweaver~\citep{rasmussen2014genome}, it is true that
we do obtain a fully resolved ARG with precise
information about recombination, but it is reasonable
to question how much information from the data is used
to support older recombinations, in particular.

It is helpful to consider the difference between fully
resolved binary trees and those that contain polytomies.

The situation is in fact directly analogous: an ARG that
contains fully resolved recombination information must have
at most parents per node.

At the very least, it is not reasonable for the \emph{data
structure} to require complete precision in order to
represent some less precise. This is a fundamental
advantage of the gARG encoding over the classical eARG
method: we are not forced to posit a precise cause (recombination
event) for every effect (difference between adjacent trees).
These differences in precision of recombination
information are illustrated in the ARGs estimated by different
methods, as we see in the next section.

>>>>

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees. The operation to remove them,
the results of which are shown in Fig.~\ref{fig-simplification}C,
therefore requires knowledge not just of the graph topology, but also of the
edge annotations. Removal of recombinant nodes can produce graph nodes with
more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node,
something which is impossible to encode in an eARG.

@@

We have seen that the local tree for a given position on
the genome can be extracted from an ARG by starting at the
leaves and following the appropriate paths through the graph.
However, this tells us little about the relationship between
adjacent trees, and what the properties of overall ensemble of
local trees are.
A recombination between two adjacent trees causes them to
differ by a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the ``subtree transfer'' operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

The effects of recombination as we move from left-to-right along
the genome in an unsimplified gARG are easy to interpret.
Consider the highlighted tree in Fig~\ref{fig-simplification}A.
In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
its parent is \textsf{k}. Since both parents are present as
unary nodes in the tree, we simply change one edge in
order to effect the transition.
% Need revising to fit new example
A more complex transition occurs when we have a regraft operation
that occurs above the current root (i.e., the current oldest
non-unary node). For example, the transition between the
second and third trees corresponds to node D switching
from the path through I to H, and we then must add branches
corresponding to passing through L and meeting at M.

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is important to note that while extracting local
trees from an ARG is straightforward, the converse
process of reconstructing precisely the same ARG from
these trees is more problematic. If the internal
nodes are consistently labelled and unary nodes are
included in the local trees (e.g., Fig~\ref{fig-simplification})
then it is easy to exactly reconstruct the ARG.
Otherwise, the problem is much more difficult and it
may not be possible to uniquely reconstruct the original ARG.
Without labels for the internal nodes, we are forced to
to reason about which subtrees remain the same and which
differ due to the SPR separating adjacent trees in order to
assign node identity. Without the internal unary nodes marking
recombinations, we must infer their existence. In
general, there will not be a unique solution to this problem,
and so there can be many possible ARGs compatible with
a given list of leaf-labelled binary trees.
This is the approach taken by the Espalier inference
method~\citep{rasmussen2022espalier}, which finds a
set of SPRs that reconcile independently inferred
trees from different regions of the genome.

\section*{Precision of ancestry information}
Ancestral recombination graphs are often characterised
as consisting of a complete record of the genetic history
of a sample. For example,
\cite{rasmussen2014genome} state that an ``ARG provides a record of all
coalescence and recombination events since the divergence of the sequences
under study'' and
according to \cite{deng2021distribution} an ARG
``provides all the information about the genealogical history of a sample,
including the locations of recombination events.'' It is worth
questioning, however, whether such comprehensive information
will always be available, and in particular whether our
representation of genealogical history should \emph{require}
such precision. Simulations will usually generate complete
information about the history of a sample, of course, but
we may not have sufficient information to infer such detail.

@@

% % JK: This is very rough, just jotting down the basic ideas,
% % will need substantial revision.
% It is not clear that such complete information will always
% (or indeed ever) be available from inferences. We may be
% able to precisely identify the location and timing
% of recent recombinations (particularly if detailed pedigree
% information is available) but beyond this, the information
% required simply isn't present. When sampling from a
% process such as the Sequentially Markov
% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
% like ARGweaver~\citep{rasmussen2014genome}, it is true that
% we do obtain a fully resolved ARG with precise
% information about recombination, but it is reasonable
% to question how much information from the data is used
% to support older recombinations, in particular.

% It is helpful to consider the difference between fully
% resolved binary trees and those that contain polytomies.

% The situation is in fact directly analogous: an ARG that
% contains fully resolved recombination information must have
% at most parents per node.

% At the very least, it is not reasonable for the \emph{data
% structure} to require complete precision in order to
% represent some less precise. This is a fundamental
% advantage of the gARG encoding over the classical eARG
% method: we are not forced to posit a precise cause (recombination
% event) for every effect (difference between adjacent trees).
% These differences in precision of recombination
% information are illustrated in the ARGs estimated by different
% methods, as we see in the next section.
